Developmental integration cannot explain major features of stomatal anatomical evolution in seed plants

### 2021-09-28 template/amnat_preamble.tex
initial amnat submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

\author{}
\date{\vspace{-2.5em}}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\author{}
\date{}

\begin{document}

\setstretch{1.7}
\maketitle

\bigskip

\textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, table 1, online appendix A (including Notes A1, table A1, table A2, and table A3).

\bigskip

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, phylogenetic comparative methods, stomata

\bigskip

\textit{Manuscript type}: Major Article

\bigskip

\textit{Word Count}: 4,704

\bigskip

\linenumbers{}
\modulolinenumbers[3]

\newpage{}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically. We synthesized data on stomatal density and length from amphistomatous leaves of 711 terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from 327 phylogenetically independent contrasts using a robust Bayesian model. Adaxial stomatal density, but not length, is evolutionarily labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal ratio. Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.

\newpage{}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The ability for traits to evolve independently of one another is a necessary prerequisite for adaptation to complex environments (\protect\hyperlink{ref-lewontin_adaptation_1978}{Lewontin 1978}). If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Adaptive evolution may be constrained if traits cannot evolve independently because they are developmentally integrated. Developmentally integrated traits have a ``disposition for covariation'' (\protect\hyperlink{ref-armbruster_integrated_2014}{Armbruster et al. 2014}), meaning that evolutionary divergence between lineages in one character will be tightly associated with divergence in another character. Allometry is a classic example of developmental integration that may constrain phenotypic evolution (reviewed in \protect\hyperlink{ref-pelabon_evolution_2014}{Pélabon et al.} (\protect\hyperlink{ref-pelabon_evolution_2014}{2014})). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection (\protect\hyperlink{ref-lande_quantitative_1979}{Lande 1979}). However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection (\protect\hyperlink{ref-hansen_is_2003}{Hansen 2003}). For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg's rule) (\protect\hyperlink{ref-berg_general_1959}{Berg 1959}, \protect\hyperlink{ref-berg_ecological_1960}{1960}; \protect\hyperlink{ref-conner_raissa_2014}{Conner and Lande 2014}; \protect\hyperlink{ref-armbruster_covariance_1999}{Armbruster et al. 1999}). In this study we are interested in quantifying the strength of developmental integration in stomatal anatomy and whether developmental integration might hamper or accelerate adaptive evolution.

Stomata are microscopic pores on the leaf or other photosynthetic surfaces of most land plants formed by a pair of guard cells. Here we limit our focus to stomatal traits on leaves within terrestrial seed plants, primarily angiosperms. The density, size, and patterning of stomata on a leaf set the maximum stomatal conductance to CO\(_2\) diffusing into a leaf and the amount of water that transpires from it (\protect\hyperlink{ref-sack_hydrology_2003}{Sack et al. 2003}; \protect\hyperlink{ref-franks_effect_2001}{Franks and Farquhar 2001}; \protect\hyperlink{ref-galmes_leaf_2013}{Galmés et al. 2013}). Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function (\protect\hyperlink{ref-de_boer_optimal_2016}{Boer et al. 2016}) and are correlated with actual stomatal function under natural conditions (\protect\hyperlink{ref-murray_consistent_2020}{Murray et al. 2020}).

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as ``inverse size-density scaling'' and ``bimodal stomatal ratio'' (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata (\protect\hyperlink{ref-weiss_untersuchungen_1865}{Weiss 1865}; \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling 2009}; \protect\hyperlink{ref-de_boer_optimal_2016}{Boer et al. 2016}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}; \protect\hyperlink{ref-liu_scaling_2021}{Liu et al. 2021}). Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species (\protect\hyperlink{ref-metcalfe_anatomy_1950}{Metcalfe and Chalk 1950}; \protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}), especially herbaceous plants from open, high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}, \protect\hyperlink{ref-muir_light_2018}{2018}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}). \protect\hyperlink{ref-muir_making_2015}{Muir} (\protect\hyperlink{ref-muir_making_2015}{2015}) described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (\emph{cf.} \protect\hyperlink{ref-peat_comparative_1994}{Peat and Fitter} (\protect\hyperlink{ref-peat_comparative_1994}{1994}) Fig 1). For both features, we limit our focus in this study to interspecific variation in mean trait values and do not seek to understand intraspecifc variation.

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in land plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, leaves with smaller stomata ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

Given the significance of stomata for plant function and global vegetation modeling, we would like to understand whether these major anatomical features are shaped primarily by adaptive or nonadaptive evolutionary forces. We briefly review adaptive hypotheses for both inverse size-density scaling and bimodal stomatal ratio. Then we discuss how developmental integration might contribute to these features, in addition to or in lieu of adaptive evolution. Stomatal size and density determine the maximum stomatal conductance to CO\(_2\) and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO\(_2\) for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO\(_2\). Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference (\protect\hyperlink{ref-zeiger_stomatal_1987}{Zeiger, Farquhar, and Cowan 1987}), incur metabolic costs (\protect\hyperlink{ref-deans_optimization_2020}{Deans et al. 2020}), and/or risk hydraulic failure (\protect\hyperlink{ref-henry_stomatal_2019}{Henry et al. 2019}). The distribution of stomata and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling} (\protect\hyperlink{ref-franks_maximum_2009}{2009}) further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

Developmental integration between stomatal size and density, mediated by meristematic cell volume, could also explain inverse size-density scaling. If the size of both guard cells and the size of pavement cells between stomata are determined by initial meristematic cell volume, then changes in cell volume early in leaf development would cause both increased stomatal size and lower density (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}). This type of developmental integration would not hinder adaptive evolution when the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by \protect\hyperlink{ref-franks_maximum_2009}{Franks and Beerling} (\protect\hyperlink{ref-franks_maximum_2009}{2009}), then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

The adaptive significance of variation in stomatal ratio is unknown, but we have some clues based on the distribution of hypo- and amphistomatous leaves. The functional significance of amphistomy is that it adds a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-gutschick_photosynthesis_1984}{Gutschick 1984}). Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate (\protect\hyperlink{ref-parkhurst_intercellular_1990}{Parkhurst and Mott 1990}). Despite this amphistomy advantage, most leaves are hypostomatous, suggesting that the benefits of amphistomy in terms of increased photosynthesis usually do not outweigh the costs of stomata on the upper surface. Amphistomy should increase photosynthesis under saturating-light conditions where CO\(_2\) supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-gibson_structure-function_1996}{Gibson 1996}; \protect\hyperlink{ref-smith_associations_1998}{W. K. Smith, Bell, and Shepherd 1998}; \protect\hyperlink{ref-jordan_using_2014}{Jordan, Carpenter, and Brodribb 2014}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}; \protect\hyperlink{ref-bucher_stomatal_2017}{Bucher et al. 2017}), especially in herbs (\protect\hyperlink{ref-muir_light_2018}{Muir 2018}). However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants (\protect\hyperlink{ref-muir_is_2019}{Muir 2019}), suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in \emph{Populus} (\protect\hyperlink{ref-mckown_association_2014}{McKown et al. 2014}, \protect\hyperlink{ref-mckown_role_2019}{2019}; \protect\hyperlink{ref-fetter_growthdefense_2021}{Fetter, Nelson, and Keller 2021}). Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits (\protect\hyperlink{ref-buckley_how_2015}{Buckley et al. 2015}). Other hypotheses about the adaptive significance of stomatal ratio are discussed in \protect\hyperlink{ref-muir_making_2015}{Muir} (\protect\hyperlink{ref-muir_making_2015}{2015}) and \protect\hyperlink{ref-drake_two_2019}{Drake et al.} (\protect\hyperlink{ref-drake_two_2019}{2019}).

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal. \protect\hyperlink{ref-muir_making_2015}{Muir} (\protect\hyperlink{ref-muir_making_2015}{2015}) derived general conditions in which bimodality arises because adaptive optima are restricted to separate regimes, but this model has not been tested. An alternative hypothesis is that stomatal traits on the ab- and adaxial surfaces are developmentally integrated because stomatal development is regulated the same way on each surface. In hypostomatous leaves, stomatal development is turned off in the adaxial surface. In amphistomatous leaves, stomatal development proceeds on both surfaces, but evolutionary changes in stomatal development affect traits on both surfaces because they are tethered by a shared developmental program. This model of developmental integration would lead to a bimodal trait distribution because leaves would either be hypostomatous (stomatal ratio equal to 0) or have similar densities on each surface (stomatal ratio approximately 0.5). To our knowledge, this hypothesis has not been put forward in the literature but came about during discussion with one of us (Muir, personal communication with EJ Edwards).

We reasoned that the developmental hypotheses for inverse size-density scaling and bimodal stomatal ratio could be tested using amphistomatous leaves. Since both leaf surfaces are formed from the same genome and meristematic cells, if developmental integration is strong we would expect similar patterns of trait covariation in the abaxial and adaxial surface (see below for specific predictions). Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can be developmentally disintegrated. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection (\protect\hyperlink{ref-barrett_sexual_2013}{Barrett and Hough 2013}). Below we reiterate the hypotheses and specific predictions for amphistomatous leaves:

\noindent {\bf{Inverse size-density scaling}}

\begin{itemize}
\item
  \emph{Hypothesis}: Meristematic cell volume mediates developmental integration between stomatal size and density
\item
  \emph{Prediction}: Amphistomatous leaves will exhibit identical size-density scaling on each surface
\end{itemize}

\noindent {\bf{Bimodal stomatal ratio}}

\begin{itemize}
\item
  \emph{Hypothesis}: Stomatal traits on both leaf surfaces are developmentally integrated because they follow the same developmental program
\item
  \emph{Prediction}: Evolutionary divergence in stomatal traits on one surface will be isometric with divergence in stomatal traits on the other surface
\end{itemize}

We tested these predictions in a phylogenetic comparative framework by compiling stomatal anatomy data from the literature for a broad range of seed plants.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in \emph{R} version 4.1.0 (\protect\hyperlink{ref-r_core_team_r:_2021}{R Core Team 2021}). Source code is publically available on GitHub (\url{https://github.com/cdmuir/stomata-independence}) and will be archived on Zenodo upon publication.

\hypertarget{data-synthesis}{%
\subsection{Data synthesis}\label{data-synthesis}}

We searched the literature for studies that measured stomatal density and stomatal size, either guard cell length or stomatal pore length, for both abaxial and adaxial leaf surfaces. In other words, we did not include studies unless they reported separate density and size values for each surface. We did not record leaf angle because it is typically not reported, but we presume that for the vast majority of taxa that the abaxial is the lower surface and the adaxial is the upper surface. This is reversed in resupinate leaves, but to the best of our knowledge, our synthesis did not include resupinate leaves. We refer to guard cell length as stomatal length and converted stomatal pore length to stomatal length assuming guard cell length is twice pore length (\protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). Table \ref{tab:traits} lists focal traits and symbols.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.20}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.52}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.28}}@{}}
\caption{\label{tab:traits}Stomatal anatomical traits with mathemtical symbol, variable string used in source code, and scientific units.}\tabularnewline
\toprule
Symbol & Variable string & Units \\
\midrule
\endfirsthead
\toprule
Symbol & Variable string & Units \\
\midrule
\endhead
\(D_\mathrm{ab}\) & \(\mathtt{abaxial\_stomatal\_density\_mm2}\) & \(\text{pores mm}^{-2}\) \\
\(D_\mathrm{ad}\) & \(\mathtt{adaxial\_stomatal\_density\_mm2}\) & \(\text{pores mm}^{-2}\) \\
\(L_\mathrm{ab}\) & \(\mathtt{abaxial\_stomatal\_length\_um}\) & \(\mu\)m \\
\(L_\mathrm{ad}\) & \(\mathtt{adaxial\_stomatal\_length\_um}\) & \(\mu\)m \\
\bottomrule
\end{longtable}

Data on stomatal anatomy are spread over a disparate literature and we have not attempted an exhaustive synthesis of amphistomatous leaf stomatal anatomy. We began our search by reviewing papers that cited key studies of amphistomy (\protect\hyperlink{ref-parkhurst_adaptive_1978}{Parkhurst 1978}; \protect\hyperlink{ref-mott_adaptive_1982}{Mott, Gibson, and O'Leary 1982}; \protect\hyperlink{ref-muir_making_2015}{Muir 2015}). We supplemented these by searching Clarivate \emph{Web of Science} for ``guard cell length'' because most studies that report guard cell length also report stomatal density, whereas the reverse is not true. We identified additional studies by reviewing the literature cited of papers we found and through haphazard discovery. The final data set contained 5104 observations of stomatal density and length from 1242 taxa and 38 primary studies (Table \ref{tab:sources}). However, many of these data were excluded if taxonomic name and phylogenetic placement could not be resolved (see below). Finally, we included some previously unpublished data. Stomatal size data were collected on grass species previously described (\protect\hyperlink{ref-pathare_increased_2020}{Pathare, Koteyeva, and Cousins 2020}). We also included a previously unpublished set of 14 amphistomatous wild tomato species (\emph{Solanum} sect. \emph{Lycopersicum} and sect. \emph{Lycopersicoides}) grown in pots under outdoor summer Mediterranean conditions previously described (\protect\hyperlink{ref-muir_unpublished_2021}{Muir, Galmés, and Conesa 2021}). We took ab- and adaxial epidermal imprints using clear nail polish of the mid-portion of the lamina away from major veins on the terminal leaflet of the youngest, fully expanded leaf from 1-5 replicates per taxon. With a brightfield light microscope, we counted stomata in three fields of view and divided by area to obtain density and measured the average guard cell length to estimate stomatal size. The data set is publicly available as an \emph{R} package \textbf{ropenstomata} (\url{https://github.com/cdmuir/ropenstomata}). It will be deposited on Dryad and archived on Zenodo upon publication.

\hypertarget{phylogeny}{%
\subsection{Phylogeny}\label{phylogeny}}

We resolved taxonomic names using the R package \textbf{taxize} version 0.9.99 (\protect\hyperlink{ref-chamberlain_taxize_2013}{Chamberlain and Szöcs 2013}). We queried taxonomic names supplied by the original study authors on 2021-08-26 from the following sources: GRIN Taxonomy for Plants (\protect\hyperlink{ref-united_states_department_of_agriculture_agricultural_research_service_germplasm_2020}{United States Department of Agriculture, Agricultural Research Service 2020}), Open Tree of Life Reference Taxonomy (\protect\hyperlink{ref-rees_automated_2017}{Rees and Cranston 2017}), The International Plant Names Index (\protect\hyperlink{ref-the_royal_botanic_gardens_international_2020}{The Royal Botanic Gardens et al. 2020}), Tropicos - Missouri Botanical Garden (\protect\hyperlink{ref-missouri_botanical_garden_tropicos_2020}{Missouri Botanical Garden 2020}). We retained the maximum scoring matched name with taxize score \(\ge 0.75\) (a score of 1 is a perfect match). In 5 ambiguous cases we manually curated names. Taxonomic name resolution reduced the data set from 1224 to 1183 taxa. Most taxa are different species, but some recognized subspecies and varieties are also included. All algorithms and choices are documented in the associated source code.

We used the R packages \textbf{taxonlookup} version 1.1.5 (\protect\hyperlink{ref-pennell_simple_2016}{Pennell, FitzJohn, and Cornwell 2016}) and \textbf{V.phylomaker} version 0.1.0 (\protect\hyperlink{ref-jin_vphylomaker_2019}{Jin and Qian 2019}) to maximize overlap between our data set and the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018} ; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}). We further resolved large (\(\ge4\) taxa) polytomies in 30 clades with sufficient sequence data using \textbf{PyPHLAWD} version 1.0 (\protect\hyperlink{ref-smith_pyphlawd_2019}{S. A. Smith and Walker 2019}) in Python 3.8.9 (Python Software Foundation, \url{https://www.python.org/}). We used sequence data from the most recent GenBank Plant and Fungal sequences database division (\protect\hyperlink{ref-ouellette_database_1997}{Ouellette and Boguski 1997}). We inferred subtree phylogenies using RAxML version 8.2.12 (\protect\hyperlink{ref-stamatakis_raxml_2014}{Stamatakis 2014}) and conducated molecular dating using the \texttt{chronos()} function in the R package \textbf{ape} version 5.5 (\protect\hyperlink{ref-paradis_ape_2019}{Paradis and Schliep 2019}) to obtain ultrametric trees. We grafted resolved, ultrametric subtrees onto the mega-tree at the polytomy nodes and rescaled to keep the mega-tree ultrametric. In some cases, resolving polytomies was not possible because there was little or no overlap between taxa in the data set and taxa with sequence data available for \textbf{PyPHLAWD}. In these cases, we randomly selected two taxa as a phylogenetially independent pair and dropped the rest. Remaining polytomies of three taxa were resolved randomly using the \texttt{multi2di()} function in \textbf{ape}. The final data set for which we had both trait and phylogenetic information contained 711 taxa (Notes A1). Seven taxa are gymnosperms; the vast majority are angiosperms.

\hypertarget{phylogenetically-independent-contrasts}{%
\subsection{Phylogenetically independent contrasts}\label{phylogenetically-independent-contrasts}}

We extracted 327 phylogenetically independent taxon pairs (Table \ref{tab:pair_div}). A fully resolved, bifurcating four-taxon phylogeny can have two basic topologies: \(\tt{((A,B),(C,D))}\) or \(\tt{((A,B),C),D))}\). Taxon pairs include all comparisons of \(\tt{A}\) with \(\tt{B}\) and \(\tt{C}\) with \(\tt{D}\) in each four-taxon clade. We extracted pairs using the \texttt{extract\_sisters()} function in R package \textbf{diverge} version 2.0.1 (\protect\hyperlink{ref-anderson_diverge_2021}{Anderson and Weir 2021}) and custom scripts (see source code). Taxon pairs are the most closely related pairs in our data set, but not usually the most closely related species in nature. For each pair we calculated phylogenetically independent contrasts (\protect\hyperlink{ref-felsenstein_phylogenies_1985}{Felsenstein 1985}) as the difference in the log\(_{10}\)-transformed trait value. Contrasts are denoted as \(\Delta \text{log(trait)}\). We log-transformed traits for normality because like many morphological and anatomical traits they are strongly right-skewed. Log-transformation also helps compare density and length, which are measured on different scales, because log-transformed values quantify proportional rather than absolute divergence. We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout seed plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.

\hypertarget{parameter-estimation}{%
\subsection{Parameter estimation}\label{parameter-estimation}}

Both hypotheses make predictions about trait covariance matrices or parameters derived from them (see next subsections). We estimated the \(4 \times 4\) covariance matrix of phylogenetically independent contrasts between log-transformed values of \(\Delta \text{log}(D_\mathrm{ab})\), \(\Delta \text{log}D_\mathrm{ad})\), \(\Delta \text{log}(L_\mathrm{ab})\), and \(\Delta \text{log}(L_\mathrm{ad})\) using a distributional multiresponse robust Bayesian approach. We denote variances as \(\text{Var}[\Delta \text{log(trait)}]\) and covariances as \(\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]\). The statistical model also accounts for differences in mean trait values between surfaces.

We used a multivariate \(t\)-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values. Exceptional trait values are common in biology and can distort estimates of central tendency and variance. The \(t\)-distribution is more robust because it has fatter tails (more kurtosis) than the Normal distribution. The \(\nu\) parameter of the \(t\)-distribution describes how fat the tails are. As \(\nu \to \infty\), the \(t\)-distribution converges to the Normal distribution. We estimated \(\nu\) from the data, meaning that if exceptional values are absent from the data, the model will be nearly equivalent to standard Normal regression.

We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g.~Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix, but not the trait mean.

We fit the model in Stan 2.27 (\protect\hyperlink{ref-stan_development_team_stan_2021}{Stan Development Team 2021}) using the R packages \textbf{brms} version 2.16.0 (\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017}, \protect\hyperlink{ref-burkner_advanced_2018}{2018}) with a \textbf{cmdstanr} version 0.4.0 backend (\protect\hyperlink{ref-gabry_cmdstanr_2021}{Gabry and Češnovar 2021}). It ran on 2 parallel chains for 1000 warm-up iterations and 1000 sampling iterations. All parameters converged (\(\hat{R} \approx 1\)) and the effective sample size from the posterior exceeded 1000 (\protect\hyperlink{ref-vehtari_rank-normalization_2021}{Vehtari et al. 2021}). We used the posterior median for point estimates and calculated uncertainty with the 95\% highest posterior density (HPD) interval from the posterior distribution.

\hypertarget{is-size-density-scaling-the-same-on-both-leaf-surfaces}{%
\subsection{Is size-density scaling the same on both leaf surfaces?}\label{is-size-density-scaling-the-same-on-both-leaf-surfaces}}

We tested the first hypothesis by estimating the covariance between divergence in stomatal length and stomatal density on each leaf surface. If size and density are developmentally integrated, we predict the covariance matrices will not be significantly different. Specifically, the 95\% HPD intervals of the difference in covariance parameters should not include 0 if:

\begin{gather}\label{eq:prediction1}
\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})] \\
\text{Var}[\Delta \text{log}(L_\text{ab})] \ne \text{Var}[\Delta \text{log}(L_\text{ad})] \\
\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]
\end{gather}

\hypertarget{do-abaxial-and-adaxial-stomatal-traits-evolve-isometrically}{%
\subsection{Do abaxial and adaxial stomatal traits evolve isometrically?}\label{do-abaxial-and-adaxial-stomatal-traits-evolve-isometrically}}

If stomatal traits on each surface are developmentally integrated then divergence in the trait on one surface should result in a 1:1 (isometric) change in the trait on the other surface. Furthermore, there should be relatively little variation away from a 1:1 relationship. Conversely, if traits can evolve independently then the change in the trait on one surface should be uncorrelated with changes on the other. We tested for isometry by estimating the standardized major axis (SMA) slope of change in the abaxial trait against change in the adaxial trait for both stomatal length and stomatal density. If change on each surface is isometric, then the HPD intervals for the slope should include 1. We used the coefficient of determination, \(r^2\), to quantify the strength of integration, where a value of 1 is complete integration and a value of 0 is complete disintegration.

\hypertarget{results}{%
\section{Results}\label{results}}

>>>>

% \documentclass[11pt]{article} % redundant
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

\date{}
